Xolotl: An Intuitive and Approachable Neuron \& Network Simulator in MATLAB

### 2018-06-11 paper/paper.tex
added network snippet
went over introduction

\def\journal{Neuroinformatics}
\def\DOI{}
\def\articleType{Technology Report}
\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Hoyland {et~al.}} 
\def\Authors{Srinivas Gorur-Shandilya\,$^{1*\dagger}$, Alec Hoyland\,$^{1\dagger}$, and Eve Marder$^{1}$}
\def\Address{$^{1}$Volen Center and Biology Department \\ Brandeis University \\ Waltham, MA 02453 \\ USA
\vspace{0.25cm}

\vspace{0.5em}\tiny{$\dagger$ These authors have equally contributed to this article.}
}
\def\corrAuthor{Srinivas Gorur-Shandilya \vspace{1mm}}
%\def\corrAddress{Marder Lab, Brandeis University, Biology Department and Volen Center for Complex Systems, Waltham, MA, USA}
\def\corrAddress{Volen Center for Complex Systems, Brandeis University, Waltham MA 02453}
\def\corrEmail{\texttt{srinivasgs@brandeis.edu}}

@@

	Information processing by neurons relies on the transmission and interaction of electrical signals that arise from the biophysics of ion channels and synapses. Electrophysiological characterization of these mechanisms has allowed for the construction of conductance-based models that can reproduce many features of neuronal and circuit behavior. However, working with conductance-based models continues to be a challenge due to their high dimensionality, hindering intuition of their dynamical features. Here, we present a neuron and network simulator built on a novel automatic type system that binds class templates written in "C++" to object-oriented code in "MATLAB". Our approach combines the speed and powerful features of "C++" code with the ease-of-use and rich libraries of scientific programming languages like "MATLAB". In our software, models are hierarchical, and components are named and searchable, permitting high-level programmatic control over all parameters. The simulator's architecture allows for the live manipulation of any parameter in any model, and for visualizing the effects of changing that parameter on model behavior. The simulator is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be easily extended to include arbitrary mechanisms and dynamics. Finally, the simulator is written in a modular fashion and has been released under a permissive free software license, enabling it to be integrated easily in third party applications.

	\tiny
	\keyFont{ \section{Keywords:} simulator, MATLAB, C++, conductance-based, neuron, network, pedagogy}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		INTRODUCTION & OVERVIEW
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Nervous systems process and transmit information using electrically excitable membranes. Conductance-based models are powerful biophysical simplification of an electrically excitable cell (\cite{hodgkinQuantitativeDescriptionMembrane1952}). Studies based on the Hodgkin-Huxley formalism now contribute significantly to mainstream research in small-circuit networks (\cite{marderTheoryMotion1995, prinzComputationalApproachesNeuronal2010, prinzInsightsModelsRhythmic2006}). Additionally, these models provide an approachable framework for understanding salient principles of neuroscience. However, several challenges remain in simulating biophysically-detailed conductance-based neuron models. First, these models can be high-dimensional with many nonlinear differential equations, with several parameters. Second, many or all equations are strongly coupled through the variables like the membrane potential, and in multi-compartment models, all membrane potentials and their equations are coupled to each other. Finally, the choice of programming language imposes tradeoffs in neuron and network simulators: simulators written in languages like "C", "C++", or "hoc" can integrate equations quickly, but often lack the ease-of-use and interoperability of those written in scientific programming languages like "Python", "Julia" or "MATLAB".

Two major approaches have dominated the design of neuron simulators. First, some simulators, such as "NEURON" (\cite{hinesNEURONSimulationEnvironment1997}) use a custom language to specify modular network components. These simulators tend to perform fast computations with little overhead, but suffer from a steep learning curve. Wrapping "NEURON" in a more approachable language like "Python" and use of graphical interfaces mitigate these drawbacks, at the cost of obfuscating the underlying algorithms and parameters (\cite{bretteSimulationNetworksSpiking2007, hinesNEURONPython2009}). In contrast, many simulators have been designed to be used entirely in popular scientific programming languages. These simulators can be easier to use, and can benefit from inter-compatibility with other tools. Programs like "DynaSim" (\cite{sherfeyDynaSimMATLABToolbox2018}), "ANNarchy" (\cite{vitayANNarchyCodeGeneration2015}), "BRIAN" (\cite{stimbergBrianSecondComing2013}) allow the user to specify models using strings of equations that are specified in the scientific programming language, that can be translated into a faster implementation language such as "C" or "C++" (\cite{stimbergEquationorientedSpecificationNeural2014}). This approach permits considerable flexibility for simulating systems of differential equations, but the syntax tends to be verbose and the hierarchical nature of conductance-based neuron models is not naturally reflected in the constructed model. Neither approach permits tools that simultaneously maintain efficiency, ease-of-use, and clarity.

To overcome these design limitations, we have developed a novel automatic type system, "cpplab", which binds "MATLAB" code to "C++" header files. This architecture automatically creates objects in "MATLAB" that represent the underlying object-oriented "C++" code, without having to re-specify these objects in "MATLAB". In this paper, we introduce "xolotl", an implementation of the "cpplab" system specialized in integrating conductance-based neuron and network models. Models can be easily constructed from network building blocks in a few lines of "MATLAB" code using a hierarchical and intuitive syntax. Since model objects in the "MATLAB" workspace are automatically linked to the "C++" specification, configuring these objects in "MATLAB" transparently configures the underlying "C++" objects. "xolotl" comes packaged with hundreds of conductances and synapses, built-in visualization functions, and a graphical user interface (GUI) for real-time manipulation of model parameters. Plotting of voltage, intracellular calcium, conductance gating functions, and time constants is provided by built-in "xolotl" methods. The GUI permits real-time tuning of any network parameters using numerical sliders in a graphical interface which displays the resultant membrane potential and intracellular calcium traces. The ease-of-use of these tools lends them to pedagogical applications and rapid exploration of toy models. This tool aims to simplify the investigation of dynamics of biophysically-realistic network models, facilitate collaborative modeling, and complement other tools being developed in the computational neuroscience community.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		DESIGN GOALS
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

"xolotl" is designed to be easy-to-use and richly featured and that is fast enough to use in everyday research. Our focus was on designing a simulator of conductance-based neurons and networks of these neurons; and simulating arbitrary dynamical systems is therefore beyond the scope of this software. The software is designed to be used from within "MATLAB", a scientific programming language popular amongst neuroscientists and engineers for pedagogy and research. Our goal is to make "xolotl" completely usable entirely within MATLAB, and all parameters of models can be be inspected and edited before running the model. Several features of the simulator are designed to be easily extensible, and the simulator appears in the "MATLAB" workspace as a first-class native object, and is fully compatible with the large library of toolboxes that a scientific programming environment like "MATLAB" provides.

@@

\paragraph{A rich library of network components.} "xolotl" comes packaged with hundreds of pre-existing components (compartments, synapses, conductances, and controllers) which are used to construct model neurons and networks. Compartments represent cylindrical sections of membrane with a single membrane potential, and can represent either entire cells or parts of cells. Synapses connect two compartments together by introducing a current in the post-synaptic neuron that depends on the voltage of the presynaptic neuron. Conductances represent populations of ion channels in a compartment that produce transmembrane currents. Controllers represent homeostatic mechanisms that integrate a Calcium-dependent signal from the compartment they are in and  change the properties of a conductance or synapse in response.

\paragraph{Object-oriented.}
"xolotl" is designed to mirror the nested and hierarchical structure of networks and neurons.  Biological neuronal networks consist of neurons that are connected to each other with synapses. Each of these neurons contains within it a set of conductances and synaptic currents that affect its electrical behavior. Similarly, a "xolotl" model can contain a set of compartments that can represent individual neurons. Each compartment can contain a set of conductances. All objects in the "xolotl" model tree (compartments, conductances, etc.) are bonafide "MATLAB" objects that have their own type, properties and associated methods. This object-oriented programming paradigm naturally represents the hierarchical structure of biological neurons and networks.

\paragraph{Automatic type system.}
When a network component is added in "MATLAB", the "xolotl" structure is updated, and "cpplab" automatically constructs and compiles the "C++" from the requisite header files. "MATLAB" treats "cpplab" constructions as fully-typed first-class objects allowing for symbolic manipulation using only the high-level programming language and graphical interfaces. "xolotl" simulations are run entirely from "C++" executables; the inputs and outputs can be represented entirely in "MATLAB".

\paragraph{Automatic compiling.}
The user experience of "xolotl" can stay entirely in "MATLAB". Transpiling and compiling of "C++" is performed automatically. Cryptographic hashing of the "xolotl" object confirms that compiling occurs only when necessary, and that the correct binary is used during integration.

@@

The focus on ease-of-use and fast simulations means some features were intentionally neglected in the design process.

@@

\paragraph{Limited numerical integration strategies.} While the exponential Euler method performs well in neuronal models (\cite{ohErrorAnalysisSpecialized2006, dayanTheoreticalNeuroscience2001}, it may be desirable to use other methods under certain conditions. "xolotl" does not currently support other integration schemes for its built-in conductances, nor does the software support error-sensitive variable step-sizes `out-of-the-box.'

\paragraph{Inefficient tools for handling large networks.} While "xolotl" can integrate large networks ($>1000$ compartments), "xolotl" uses string-based comprehension for labeling compartments which is suited to descriptive naming, but prohibits vector operations over compartments.

\paragraph{New mechanisms require new \texttt{C++} code.} Adding new network components also requires writing some "C++" code. Adding a new conductance in the Hodgkin-Huxley formalism (\cite{hodgkinMeasurementCurrentvoltageRelations1952, dayanTheoreticalNeuroscience2001}), requires creating a new "C++" header file, though this is generally trivial. Implementing a new integration scheme or "cpplab" class requires much more in-depth knowledge of "C++" and modification of the core code.

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_HH}
	\caption{\texttt{xolotl} can quickly set up and simulate conductance-based models. Schematic representation of a electrotonically-neuron with three populations of ion channels (colored rectangles). In the simulator, the soma is represented using ab object of class $compartment$ and the populations of ion channels are represented by $conductance$ objects contained within the compartment. Code snippet used to set up this neuron model, inject current, integrate and plot the voltage, and show activation functions demonstrates how various several steps can be run in a few lines of code. (A) Simulated voltage trace of a Hodgkin-Huxley model with three conductances and 0.2 nA of injected current. Colors indicate the dominant current (gold is fast sodium (NaV), blue is delayed rectifier (Kd), red is Leak). (C) f-I curve of this neuron. (D-E) Steady-state gating functions for activation (m) and inactivation (h) gating variables. (F-I) Voltage-dependence of time constants for activation (m) and inactivation (h) gating variables}
	\label{fig:figurehh}
\end{figure}

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		GRAPHICAL INTERFACE
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_clamp}
	\caption{Simulating a voltage-clamp experiment. (A) Cartoon of a cell with delayed rectifier potassium conductance (\cite{liuModelNeuronActivitydependent1998} with experimentally-fixed voltage. (B) Structure of \texttt{xolotl} object in A. (C) Code snippet depicting integration under voltage clamp. (D-E) Current response to steps in voltage from a holding potential of $V_m = -60$ mV. (F) Current-voltage relation of the steady-state current ($t = 400$ ms) indicating a reversal potential of $E = -80$ mV and no inactivation. (G) Conductance-voltage relation at steady-state takes the form of a sigmoid. (H) Sigmoids $m$ fit to the model as $m^n$ data indicating that $n=4$ is the best fit. (I) Goodness of fit vs. exponent $n$, suggesting $n=4$ as the best fit to the data.}
	\label{fig:figureclamp}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_network}
	\caption{Simulating a network of conductance-based model neurons. (A) Diagram of a network model of the pyloric rhythm in the crustacean stomatogastric ganglion (Prinz \textit{et al.} 2004). (B) Each neuron is modeled as a single compartment with 7-8 intrinsic conductances and 1-3 post-synaptic conductances. (C) \texttt{xolotl} implements conductances as fields of compartments and synapses as connections between compartments. (D-F) Simulated voltage trace of a model network for the three compartments. (G) Time series of synaptic currents in the simulated network can be obtained from the integration.}
	\label{fig:figurenetwork}
\end{figure}

>>>>

\def\journal{Neuroinformatics}
\def\DOI{}
\def\articleType{Technology Report}
\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Gorur-Shandilya, Hoyland \& Marder} 
\def\Authors{Srinivas Gorur-Shandilya\,$^{1*\dagger}$, Alec Hoyland\,$^{1\dagger}$, and Eve Marder$^{1}$}
\def\Address{$^{1}$Volen Center for Complex Systems and Biology Department \\ Brandeis University \\ Waltham, MA 02453 \\ USA
\vspace{0.25cm}

\vspace{0.5em}\tiny{$\dagger$ These authors have equally contributed to this article.}
}
\def\corrAuthor{Srinivas Gorur-Shandilya }
%\def\corrAddress{Marder Lab, Brandeis University, Biology Department and Volen Center for Complex Systems, Waltham, MA, USA}
\def\corrAddress{Volen Center for Complex Systems, Brandeis University, Waltham MA 02453}
\def\corrEmail{\texttt{srinivasgs@brandeis.edu}}

@@

	Information processing by neurons relies on the generation and transmission of electrical signals that arise from the biophysics of ion channels, synapses and other voltage-coupled components. Electrophysiological characterization of these mechanisms has allowed for the construction of conductance-based models that can reproduce several features of neuronal and circuit behavior. However, working with conductance-based models continues to be a challenge due to their high dimensionality and large number of parameters, hindering intuition of their dynamical features. Here, we present a neuron and network simulator built on a novel automatic type system that binds class templates written in "C++" to object-oriented code in "MATLAB". Our approach unites the speed and power of "C++" code with the ease-of-use and rich libraries of scientific programming languages like "MATLAB". In our software, models are hierarchical, with components that are named and searchable, permitting high-level programmatic control over all parameters. The simulator's architecture allows for the live manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. The simulator is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be easily extended to include arbitrary mechanisms and dynamics. Finally, the simulator is written in a modular fashion and has been released under a permissive free software license, enabling it to be integrated easily into the workflow of many researchers.

	\tiny
	\keyFont{ \section{Keywords:} simulator, code:MATLAB, code:C++, conductance-based, neuron, network, pedagogy}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  								%
%		INTRODUCTION 			%
% 		& OVERVIEW   			%
%								%
%								%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Nervous systems process and transmit information using electrically excitable membranes. Conductance-based models are a powerful biophysical simplification of an excitable compartment in a neuron (\cite{hodgkinQuantitativeDescriptionMembrane1952}). Studies based on the Hodgkin-Huxley formalism now contribute significantly to mainstream research in small circuits (\cite{marderTheoryMotion1995, prinzComputationalApproachesNeuronal2010, prinzInsightsModelsRhythmic2006}). These models provide an approachable framework for understanding many salient principles of neuroscience, since they explicitly model cell membranes and ion channels as electrical components in a circuit. However, several challenges remain in working with biophysically-detailed conductance-based neuron models. First, these models can be high-dimensional with many nonlinear differential equations, with several parameters. Second, many or all equations in these models can be strongly coupled through variables like the membrane potential.  In multi-compartment models, membrane potentials in every compartment are coupled to each other. Finally, the choice of programming language used to implement the model imposes tradeoffs in designing and using neuron and network simulators: simulators written in languages like "C++", "FORTRAN" or "hoc" can integrate equations quickly, but often lack the ease-of-use and interoperability of those written in scientific programming languages like "Python", "Julia" or "MATLAB".

Two major approaches have dominated the design of neuron simulators. First, some simulators, such as "NEURON" (\cite{hinesNEURONSimulationEnvironment1997}) are written in a fast compiled language like "C" and allow for the construction and simulation of neuron models using object-oriented paradigms. These simulators tend to perform fast computations with little overhead, but suffer from a steep learning curve. Wrapping these simulators in a more approachable language like "Python" and the use of graphical user interfaces (GUIs) mitigate these drawbacks, at the cost of obfuscating the underlying algorithms and parameters (\cite{bretteSimulationNetworksSpiking2007, hinesNEURONPython2009}). Another approach is to design simulators that can be used entirely in popular scientific programming languages. These simulators can be easier to use, and can benefit from inter-compatibility with other tools. Programs like "DynaSim" (\cite{sherfeyDynaSimMATLABToolbox2018}), "ANNarchy" (\cite{vitayANNarchyCodeGeneration2015}), "BRIAN" (\cite{stimbergBrianSecondComing2013}) allow the user to specify models using strings of equations that are specified in the scientific programming language, that can then be translated into a faster implementation language such as "C" or "C++" (\cite{stimbergEquationorientedSpecificationNeural2014}). This approach permits considerable flexibility for simulating systems of differential equations. Since models need to be translated between the two languages, the hierarchical nature of neuron models is not naturally encapsualted by these tools, and the syntax can be verbose. Neither approach facilitates the creation of tools that simultaneously maintain efficiency, ease-of-use, and clarity.

To overcome these design limitations, we have developed a novel automatic type system, "cpplab", which binds "MATLAB" code to "C++" header files. This architecture automatically creates objects in "MATLAB" that represent the underlying object-oriented "C++" code, without having to re-specify these objects in "MATLAB". In this paper, we introduce "xolotl", an implementation of the "cpplab" system specialized in integrating conductance-based neuron and network models. Models can be easily constructed from network building blocks in a few lines of "MATLAB" code using a hierarchical and intuitive syntax. Since models in the "MATLAB" workspace are automatically linked to models in the "C++" implementation, configuring these objects in "MATLAB" transparently configures the underlying "C++" objects. "xolotl" comes packaged with hundreds of conductances and synapses, built-in visualization functions, and a graphical user interface (GUI) for real-time manipulation of model parameters. Built-in methods allow the user to easily generate plots of voltage and intracellular calcium time traces, and activation functions of constituent conductances. The GUI permits real-time manipulation of any network parameter and an immediate visualization of the effect of that parameter on arbitrary dynamical features of the model. The ease-of-use of these tools lends them to pedagogical applications and the rapid prototyping of models. This tool aims to simplify the investigation of the dynamics of conductance-based network and neuron models, facilitate collaborative modeling, and complement other tools being developed in the computational neuroscience community.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		DESIGN GOALS
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

"xolotl" is designed to be easy-to-use and richly featured while being fast enough to use in everyday research. Our focus was on designing a simulator of conductance-based neurons and networks of these neurons; simulating arbitrary dynamical systems is therefore beyond the scope of this software. Specifically, the software was designed to simulate models of the form

\[ C_{i}\frac{dV_{i}}{dt}=-\sum_{j}I_{j} \]

where \( C_{i} \) and  \( V_{i} \) are the capacitance and membrane potential of compartment  \( i \) and \( I_{j} \) is the current due to ion channel population \( j \) given by

\[ I_{j}=\bar{g}_{j}m_{j}^{p}h_{j}^{q}(V-E_{j})A_{i} \]

Here, \( \bar{g}_{j} \) is the maximal conductance, and \( E_{j} \) is the reversal potential of the ion channel population \( j \). \( A_{i} \) is the area of compartment \( i\) that contains these ion channels.  \( m_{j} \) and \( h_{j} \) are activation and inactivation variables that change according to

\begin{eqnarray*}
\tau_{m}\frac{dm}{dt}=m_{\infty}-m & \mathrm{and} & \tau_{h}\frac{dh}{dt}=h_{\infty}-h
\end{eqnarray*}

Typically, \( \tau_{m} \), \( \tau_{h} \), \( m_{\infty} \), and \( h_{\infty} \) are functions of the membrane potential \( V_{i} \). The software uses integration schemes that have been specifically developed to solve equations of this form (cite D\&A, Hines, others), though other schemes can be used.

This software is designed to be used from within "MATLAB", a scientific programming language common amongst neuroscientists and engineers for pedagogy and research. Our goal was to make "xolotl" completely usable entirely from within MATLAB, and fully scriptable so that the software can be used as a component of other packages and tools. All parameters of models and activation functions of any channel can be be accessed before running the model. Finally, we designed several features of the simulator to be easily extensible: adding custom conductances or synapse types is possible by changing a few lines in a template file. Models created using this simulator appear in the "MATLAB" workspace as native objects, and are fully compatible with the large library of toolboxes that "MATLAB" provides.

@@

\paragraph{A rich library of network components.} "xolotl" comes packaged with hundreds of pre-existing components (compartments, synapses, conductances, and controllers) that can be used as building blocks to construct model neurons and networks. Compartments represent sections of membrane with a single membrane potential, and can represent either entire cells or parts of cells. Conductances represent populations of ion channels in a compartment that produce transmembrane currents. Synapses connect two compartments together by introducing a current in the post-synaptic neuron that can depend on the voltage of the presynaptic neuron. Controllers represent homeostatic mechanisms that integrate a Calcium-dependent signal from the compartment they are in and  change the properties of a conductance or synapse in response.

\paragraph{Object-oriented.}
"xolotl" is designed to mirror the nested and hierarchical structure of networks and neurons. Biological neuronal networks are made up of neurons that are connected to each other with synapses. Each of these neurons contains within it a set of conductances and synaptic currents that contribute to its electrical behavior. These conductances can be controlled by a feedback system that can homeostatically regulate their properties. Similarly, a "xolotl" model can contain a set of compartments that can represent individual neurons. Each compartment can contain a arbitrary set of conductances, that may be regulated by a set of controllers. All objects in the "xolotl" model tree (compartments, conductances, etc.) are bonafide "MATLAB" objects that have their own type, properties and associated methods. This object-oriented programming paradigm naturally represents the hierarchical structure of biological networks that contain neurons that contain populations of ion channels, and makes constructing and integrating these models easier.

\paragraph{Automatic type system.}
Our approach to circumvent the tradeoff between high-performance but hard to use languages like "C++" and richly featured but slow languages like "MATLAB" was to construct an automatic type system that links object oriented code in "C++" to object oriented code in "MATLAB". This architecture lets us construct the core of the simulator in "C++", leveraging features of "C++" like pointers that do not exist in "MATLAB". A rudimentary way to make this "C++" code useable in "MATLAB" would be to re-write that code in "MATLAB" so that "MATLAB" objects can be bound their "C++" implementations. However, this approach is cumbersome, and prone to errors, since the same code needs to be written twice, in two languages. Instead, our automatic type system creates objects in "MATLAB" on-the-fly  from "C++" class specifications, obviating the need to rewrite code in "MATLAB" while preserving a tight coupling between objects in the "MATLAB" workspace and their "C++" implementation.

\paragraph{Automatic hashing and compiling.}
Since every model requires a compiled binary to run, a potential stumbling block is the problem of unambiguously linking a model to a binary executable. "xolotl" solves this problem by hashing the "C++" header files of every component in the model recursively, allowing a model, no matter how complex, to be compactly represented by a short alphanumeric identifier. Compiled binaries are named with this hash, ensuring both that the correct binary is run to integrate the model, and that compilation occurs only as needed. This powerful feature enables the user experience to remain entirely within the "MATLAB" workspace, with compilation and selection of the correct binary occurring silently in the background.

@@

Our focus on ease-of-use and fast simulations means some features were intentionally neglected in the design process.

@@

\paragraph{Limited numerical integration strategies.} Most components in the software are integrated using the exponential Euler method, which has been shown to perform well in neuronal models (\cite{ohErrorAnalysisSpecialized2006, dayanTheoreticalNeuroscience2001}. However, it may be desirable to use other methods under certain conditions. It is possible to introduce new network components that implement other integration schemes, but that requires writing new "C++"  code, and is still restricted to integration schemes that assume a fixed step size.

\paragraph{Inefficient tools for handling large networks.} Since every compartment and component in a "xolotl" model is $named$, rather than numbered, it is more suited towards simulating small but complex networks, rather than large but simple networks. While the software can integrate large networks ($>1000$ compartments), other tools are presumably more suited to this task.

\paragraph{New mechanisms require new \texttt{C++} code.} Adding new network components requires writing new "C++" code. Adding a new conductance in the Hodgkin-Huxley formalism (\cite{hodgkinMeasurementCurrentvoltageRelations1952, dayanTheoreticalNeuroscience2001}), requires creating a new "C++" header file, though this is generally trivial. Implementing a new integration scheme or component type requires much more in-depth knowledge of the underlying "C++" core code.

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_HH}
	\caption{\texttt{xolotl} can quickly set up and simulate conductance-based models. Schematic representation of a electrotonically-compact neuron with three populations of ion channels (colored rectangles). In the simulator, the soma is represented using an object of class $compartment$ and the populations of ion channels are represented by $conductance$ objects contained within the compartment object. Code snippet used to set up this neuron model, inject current, integrate and plot the voltage, and show activation functions demonstrates how various several steps can be run in a few lines of code. (A) Simulated voltage trace of a Hodgkin-Huxley model with three conductances and 0.2 nA of injected current. Colors indicate the dominant current (gold is fast sodium (NaV), blue is delayed rectifier (Kd), red is Leak). (C) f-I curve of this neuron. (D-E) Steady-state gating functions for activation (m) and inactivation (h) gating variables. (F-I) Voltage-dependence of time constants for activation (m) and inactivation (h) gating variables}
	\label{fig:figurehh}
\end{figure}

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_clamp}
	\caption{Simulating a voltage-clamp experiment. (A) Cartoon of a cell with delayed rectifier potassium conductance (\cite{liuModelNeuronActivitydependent1998} with experimentally-fixed voltage. (B) Structure of \texttt{xolotl} object in A. (C) Code snippet depicting integration under voltage clamp. (D-E) Current response to steps in voltage from a holding potential of $V_m = -60$ mV. (F) Current-voltage relation of the steady-state current ($t = 400$ ms) indicating a reversal potential of $E = -80$ mV and no inactivation. (G) Conductance-voltage relation at steady-state takes the form of a sigmoid. (H) Sigmoids $m$ fit to the model as $m^n$ data indicating that $n=4$ is the best fit. (I) Goodness of fit vs. exponent $n$, suggesting $n=4$ as the best fit to the data.}
	\label{fig:figureclamp}
\end{figure}

@@

\begin{figure}[!htb]
	\centering
	\includegraphics[width=1.0\linewidth]{gfx/figure_network}
	\caption{Simulating a network of conductance-based model neurons. (A) Diagram of a network model of the pyloric rhythm in the crustacean stomatogastric ganglion (Prinz \textit{et al.} 2004). (B) Each neuron is modeled as a single compartment with 7-8 intrinsic conductances and 1-3 post-synaptic conductances. (C) \texttt{xolotl} implements conductances as fields of compartments and synapses as connections between compartments. (D-F) Simulated voltage trace of a model network for the three compartments. (G) Time series of synaptic currents in the simulated network can be obtained from the integration.}
	\label{fig:figurenetwork}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		GRAPHICAL INTERFACE
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@
